---
title: "Response time analyses"
output: rmarkdown::html_vignette
message: false
bibliography: "references.bib"
csl: "apa.csl"
vignette: >
  %\VignetteIndexEntry{Response time analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the analyses of
participants' total response times on the reasoning problems. Only Bayesian
analyses were reported in the manuscript for brevity, but equivalent frequentist analyses
were also conducted to test the convergence of the two approaches on the same
results.

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready and clustered data (see
`vignette("preparing_data")` and `vignette("osivq_clusters")` for details). For
RT analyses, we removed incorrect trials and trials with suspicious RT patterns.
These filtering steps were gathered in the `filter_trials_on_rt()` function,
which also provides a short summary of the process.

```{r}
#| label: data-rt

df_rt <-
  get_clustered_data("experiment") |>
  filter_trials_on_rt(verbose = TRUE) |> 
  dplyr::select(id, group_4:strategy_group, problem, category, rt_total)

dplyr::glimpse(df_rt)
```

# Method

We fitted hierarchical linear models (said "Multilevel Models" in Bayesian or
"Generalised Mixed Models in frequentist) with skewed families of distributions 
to account for the typically skewed distributions of RTs, using the brms package for Bayesian models [@burknerBrmsPackageBayesian2017] or the glmmTMB package for frequentist models [@glmmTMB2025]. The
models included a grouping variable (VVIQ groups, OSIVQ clusters), Category
(visual, spatial, or control) along with their two-way interactions as fixed
categorical predictors. Varying slopes and intercepts (“random effects”) have
been added for each participant by category along with an intercept for each
problem.

Let's break this down.

## Grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_rt` data frame:

-   `group` is the 4-group VVIQ classification with an "aphantasia" group (VVIQ
    = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group (VVIQ
    $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It was
    not used in the analyses because the hyperphantasia group was too small (N =
    4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

-   `strategy_group` is a classification based on the self-reported strategies
    used by the participants to solve the problems. It focuses on whether
    participants used a visual mental imagery strategy and contains two groups:
    "Visual strategy user" and "No visual strategy". It is not reported in the
    manuscript because it was suggested to us after submission by colleagues at
    a conference.

The same modelling pipeline was therefore applied four times, once for each of
the last four grouping variables.

## Model formula

We created a small `build_formula()` helper function to write the formula
easily, as we used the same model structure a lot of times:

```{r}
#| label: build-formula

build_formula("rt_total", "group_2")
 
build_formula("rt_total", "cluster")
```

## Bayesian models

```{r}
#| label: bayesian-setup
#| message: false
#| warning: false

if(!requireNamespace("cmdstanr", quietly = TRUE)) {
  install.packages(
    "cmdstanr",
    repos = c('https://stan-dev.r-universe.dev', getOption("repos"))
  )
}
options("marginaleffects_safe" = FALSE)
options("marginaleffects_posterior_interval" = "hdi")
draws <- seq(1, 2000, 1) # To limit draws that will be used for marginaleffects

prior_rt <- c(
  brms::prior(normal(3, 0.05), class = "Intercept"),
  brms::prior(normal(0, 1), class = "b"),
  brms::prior(exponential(10), class = "sd"),
  brms::prior(exponential(10), class = "sigma")
  )
```

```{r}
#| label: test_rt_prior
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Prior predictive distributions.

mb_rt_prior <-
  fit_brms_model(
    formula = build_formula("rt_total", "group_2"),
    data    = df_rt,
    family  = brms::shifted_lognormal(),
    prior   = prior_rt,
    sample_prior = "only",
    iterations = 5000,
    file = "models/m_rt_prior.rds"
  )

performance::check_predictions(mb_rt_prior) |> 
  plot() + 
  ggplot2::scale_x_continuous(limits = c(0, 60)) + 
  theme_pdf(base_size = 12)
```

## Frequentist models

The models were fitted using the `glmmTMB::glmmTMB()` function (and a little
`set_ranef_prior()` helper function to set a weakly informative prior on the
random effects, which helps with convergence and singularity issues). After
model fit, we checked for potential singularity issues and model performance
using the `get_singularity()` and `get_performance()` functions created for the
occasion (which are convenient wrappers around the *performance* package).

Finally, we tested our hypotheses with marginal contrasts. This task was
performed with the `report_contrast()` function, which is a wrapper around
several functions from the *emmeans* package. We used it to compute response
time contrasts between groups, contrasts between categories within each group,
and interaction contrasts (differences in category contrasts between groups).

Here we go!

# Results

## VVIQ 2 groups

### Bayesian

```{r}
#| label: bayesian-vviq-2-groups
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_rt_vviq_2 <-
  fit_brms_model(
    formula = build_formula("rt_total", "group_2"),
    data    = df_rt,
    family  = brms::shifted_lognormal(),
    prior   = prior_rt,
    file    = "models/m_rt_vviq_2.rds"
  )

# Singularity check
mb_rt_vviq_2 |> get_singularity()
# Model performance indices
mb_rt_vviq_2  |> 
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_rt_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_rt_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_2" = "pairwise"), draw_ids = seq(1, 4000, 2)
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_rt_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"), by = "group_2", 
    draw_ids = draws
  ) |>
  report_rope(group_2, contrast) |> knitr::kable()
# Interaction contrasts
mb_rt_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "group_2",
    hypothesis = ~revpairwise, # for the interaction
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-2-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_rt_vviq_2 <-
  glmmTMB::glmmTMB(
    data    = df_rt,
    formula = build_formula("rt_total", "group_2"),
    family  = Gamma(link = "identity"),
    prior   = set_ranef_prior(70)
  )

# Singularity check
mf_rt_vviq_2 |> get_singularity()
# Model performance indices
mf_rt_vviq_2 |> 
  get_performance(metrics = c("AICc", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_rt_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mf_rt_vviq_2 |> report_contrast(~ group_2) |> knitr::kable()
# Category contrasts within groups
mf_rt_vviq_2 |> report_contrast(~ category | group_2) |> knitr::kable()
# Interaction contrasts
mf_rt_vviq_2 |> report_contrast(~ category * group_2, interaction = TRUE) |>
  knitr::kable()
```

## VVIQ 3 groups

### Bayesian

```{r}
#| label: bayesian-vviq-3-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_rt_vviq_3 <-
  fit_brms_model(
    formula = build_formula("rt_total", "group_3"),
    data    = df_rt,
    family  = brms::shifted_lognormal(),
    prior   = prior_rt,
    file    = "models/m_rt_vviq_3.rds"
  )

# Singularity check
mb_rt_vviq_3 |> get_singularity()
# Model performance indices
mb_rt_vviq_3  |> 
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_rt_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_rt_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_3" = "revpairwise"),
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_rt_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "group_3",
    draw_ids = draws
  ) |>
  report_rope(group_3, contrast) |> knitr::kable()
# Interaction contrasts
mb_rt_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "group_3",
    hypothesis = ~revpairwise, # for the interaction
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-3-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_rt_vviq_3 <-
  glmmTMB::glmmTMB(
    data    = df_rt,
    formula = build_formula("rt_total", "group_3"),
    family  = Gamma(link = "identity"),
    prior   = set_ranef_prior(70)
  )

# Singularity check
mf_rt_vviq_3 |> get_singularity()
# Model performance indices
mf_rt_vviq_3 |> 
  get_performance(metrics = c("AICc", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_rt_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_rt_vviq_3 |> report_contrast(~ group_3) |> knitr::kable()
mf_rt_vviq_3 |> report_contrast(~ category | group_3) |> knitr::kable()
mf_rt_vviq_3 |> report_contrast(~ category * group_3, interaction = TRUE) |>
  knitr::kable()
```

## OSIVQ 3 clusters

### Bayesian

```{r}
#| label: bayesian-osivq-3-clusters
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_rt_osivq <-
  fit_brms_model(
    formula = build_formula("rt_total", "cluster"),
    data    = df_rt,
    family  = brms::shifted_lognormal(),
    prior   = prior_rt,
    file    = "models/m_rt_cluster.rds"
  )

# Singularity check
mb_rt_osivq |> get_singularity()
# Model performance indices
mb_rt_osivq  |> 
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_rt_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_rt_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("cluster" = "pairwise"),
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_rt_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "cluster",
    draw_ids = draws
  ) |>
  report_rope(cluster, contrast) |> knitr::kable()
# Interaction contrasts
mb_rt_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "cluster",
    hypothesis = ~revpairwise, # for the interaction
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```


### Frequentist

```{r}
#| label: frequentist-osivq-3-clusters
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_rt_osivq <-
  glmmTMB::glmmTMB(
    data    = df_rt,
    formula = build_formula("rt_total", "cluster"),
    family  = Gamma(link = "identity"),
    prior   = set_ranef_prior(70)
  )

# Singularity check
mf_rt_osivq |> get_singularity()
# Model performance indices
mf_rt_osivq |>
  get_performance(metrics = c("AICc", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_rt_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_rt_osivq |> report_contrast(~ cluster) |> knitr::kable()
mf_rt_osivq |> report_contrast(~ category | cluster) |> knitr::kable()
mf_rt_osivq |> report_contrast(~ category * cluster, interaction = TRUE) |>
  knitr::kable()
```

## Strategy groups

### Bayesian

```{r}
#| label: bayesian-strategy-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_rt_strat <-
  fit_brms_model(
    formula = build_formula("rt_total", "strategy_group"),
    data    = df_rt,
    family  = brms::shifted_lognormal(),
    prior   = prior_rt,
    file    = "models/m_rt_strat.rds"
  )

# Singularity check
mb_rt_strat |> get_singularity()
# Model performance indices
mb_rt_strat  |> 
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_rt_strat |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_rt_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy_group" = "pairwise"),
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_rt_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "strategy_group",
    draw_ids = draws
  ) |>
  report_rope(strategy_group, contrast) |> knitr::kable()
# Interaction contrasts
mb_rt_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "strategy_group",
    hypothesis = ~revpairwise, # for the interaction
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-strategy-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_rt_strat <-
  glmmTMB::glmmTMB(
    data    = df_rt,
    formula = build_formula("rt_total", "strategy_group"),
    family  = Gamma(link = "identity"),
    prior   = set_ranef_prior(70)
  )

# Singularity check
mf_rt_strat |> get_singularity()
# Model performance indices
mf_rt_strat |>
  get_performance(metrics = c("AICc", "R2", "RMSE")) |> 
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_rt_strat |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_rt_strat |> report_contrast(~ strategy_group) |> knitr::kable()
mf_rt_strat |> report_contrast(~ category | strategy_group) |> knitr::kable()
mf_rt_strat |> report_contrast(~ category * strategy_group, interaction = TRUE) |>
  knitr::kable()
```

# Visualisation

The figures showing the distribution of response times displayed in the
manuscript were created with the `plot_superb_raincloud()` function from the
package, which wraps functions from the *ggplot2* and *superb* packages to
create nice visualisations easily. A little `add_significance()` helper function
was also created to add significance stars to the plots.

```{r}
#| label: plot-rt
#| fig.width: 10
#| fig.height: 5
#| fig.alt: >
#|  A jitter plot showing the mean RTs in each category for several
#|  different grouping variables.

library(patchwork)
library(superb)

pr1 <-
  plot_superb_raincloud(
    df_rt, rt_total, group_2,
    title = "VVIQ 2 groups", y_title = "Mean total RT (s)",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel_x = 1.2,
    exp_add_right = 0.7
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 2,
      y_star = 30,
      stars  = "**",
      x_line = .data$x_star - 0.14,
      x_line_end = .data$x_star + 0.14,
      y_line = 29.5
    )
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 2.07,
      y_star = 28,
      stars  = "*",
      x_line = .data$x_star - 0.07,
      x_line_end = .data$x_star + 0.07,
      y_line = 27.5
    )
  )

pr2 <-
  plot_superb_raincloud(
    df_rt, rt_total, group_3,
    title = "VVIQ 3 groups", y_title = "Mean total RT (s)",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel_x = 1.2,
    exp_add_right = 0.7
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 30,
      stars  = "**",
      x_line = .data$x_star - 0.14,
      x_line_end = .data$x_star + 0.14,
      y_line = 29.5
    )
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3.07,
      y_star = 28,
      stars  = "*",
      x_line = .data$x_star - 0.07,
      x_line_end = .data$x_star + 0.07,
      y_line = 27.5
    )
  )

pr3 <-
  plot_superb_raincloud(
    df_rt, rt_total, cluster,
    title = "OSIVQ clusters", y_title = "Mean total RT (s)",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel_x = 1.2,
    exp_add_right = 0.7
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 30,
      stars  = "***",
      x_line = .data$x_star - 0.14,
      x_line_end = .data$x_star + 0.14,
      y_line = 29.5
    )
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3.07,
      y_star = 28,
      stars  = "*",
      x_line = .data$x_star - 0.07,
      x_line_end = .data$x_star + 0.07,
      y_line = 27.5
    )
  )

pr <- pr1 + pr2 + pr3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(pr)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
