---
title: "Exploring RTs per phase (non-linear modelling)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring RTs per phase (non-linear modelling)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the exploratory non-linear modelling
conducted to analyse the evolution of participants' response times over the
course of the trial phases. As stated in the manuscript:

> In addition to the total response times per trial usually analysed in VIIE
> studies, we collected response time data for each trial phase (three premises
> and the conclusion). We explored whether the VIIE could be specific to certain
> phases, or whether it was a difference in the dynamics of reasoning across
> trial phases rather than an overall difference in speed, by modelling response
> times throughout a trial with non-linear models.

The non-linear modelling used here was heavily inspired by [this tutorial by
Ping Hei Yeung](https://yeungpinghei.github.io/).

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready and clustered data (see
`vignette("preparing_data")` and `vignette("osivq_clusters")` for details)
and filter trials based on RTs (see `vignette("analysing_rt")` for details).

```{r}
#| label: get-data

df_rt <-
  get_clustered_data("experiment") |>
  filter_trials_on_rt() |> 
  dplyr::select(id:conclusion_rt)

dplyr::glimpse(df_rt)
```

By default, the `df_rt` data frame has one variable (one column) containing the
response times of each participant for each trial phase. We
needed to have all of these phases (premises and conclusion) in a single "phase"
variable along with the associated "rt" variable for modelling, so we gathered
these four columns into two long columns by "pivoting" the RT data in a long
format (reducing the number of columns and increasing the number of rows). This
operation is performed by the `pivot_phases_longer()` function.

We also needed to create "interaction" variables between the grouping variables
and the categories of the problems (visual, spatial, control) using the
`base::interaction()` function because *mgcv* models used for non-linear
modelling do not support interaction terms natively.

```{r}
#| label: pivot-terms

df_rt_long <-
  pivot_phases_longer(df_rt) |>
  dplyr::mutate(
    group_2_category = interaction(group_2, category),
    group_3_category = interaction(group_3, category),
    cluster_category = interaction(cluster, category)
  )

dplyr::glimpse(df_rt_long)
```

# Method

We fitted generalised additive models using the mgcv package (Wood, 2011) to
predict response times with a grouping variable (VVIQ groups, OSIVQ clusters),
Category (visual, spatial or control) and their interaction as fixed
categorical predictors, as well as “smooth” (non-linear) terms that capture
the evolution of RTs across trial terms (premise 1/2/3 and conclusion) for
each grouping and category, each problem for each grouping, and each
participant for each category (the latter two being equivalent to “random
effects” in mixed models).

Let's break this down.

## The grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_rt_long` data frame:

-   `group_4` is the 4-group VVIQ classification with an "aphantasia" group (VVIQ
    = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group (VVIQ
    $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It was
    not used in the analyses because the hyperphantasia group was too small (N =
    4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

The same modelling pipeline was therefore applied three times, once for each of
the last three grouping variables.

## Bayesian models

```{r}
#| label: bayesian-setup
#| message: false
#| warning: false

if(!requireNamespace("cmdstanr", quietly = TRUE)) {
  install.packages(
    "cmdstanr",
    repos = c('https://stan-dev.r-universe.dev', getOption("repos"))
  )
}
options("marginaleffects_safe" = FALSE)
draws <- seq(1, 2000, 1) # To limit draws that will be used for marginaleffects
```

## Frequentist models

The models were fitted using the `mgcv::bam()` function. After model fit, we
tested group and category differences with marginal contrasts. This task was
performed with the `get_contrast()` function, which is a wrapper around several
functions from the *emmeans* package.

These models took a long time to fit and needed to be saved for ease of reuse,
but model fits were too heavy to be kept as package data. Thus, two sub-products
of the models were saved as package data instead:

-   Contrast analyses obtained with `get_contrast()` for the statistical tests.
    The analyses for the three grouping variables have been saved in the
    `nl_contrasts` list.

-   Model predictions obtained with `modelbased::estimate_relation()` for the
    data used for visualisation. The predictions for the three grouping
    variables have been saved in the `nl_predictions` list.

The code to fit the models and get the contrasts is provided below but does not
run.

# Results

## VVIQ 2 groups

### Bayesian

```{r}
#| label: bayesian-vviq-2-groups
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

prior_nl <- c(
  brms::prior(normal(0, 1), class = "b"),
  brms::prior(exponential(1), class = "sds"),
  brms::prior(exponential(1), class = "sigma")
)

mb_nl_vviq_2 <-
  fit_brms_model(
    formula = rt ~ 
      group_2 * category + 
      s(phase, by = interaction(group_2, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_vviq_2.rds"
  )

mb_nl_vviq_3 <-
  fit_brms_model(
    formula = rt ~ 
      group_3 * category + 
      s(phase, by = interaction(group_3, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_vviq_3.rds"
  )

mb_nl_osivq <-
  fit_brms_model(
    formula = rt ~ 
      cluster * category + 
      s(phase, by = interaction(cluster, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_osivq.rds"
  )
```

```{r}
marginaleffects::avg_comparisons(
  mb_nl_vviq_2, 
  variables = list("category" = "pairwise"), 
  by = c("group_2", "phase"), 
  draw_ids = 1:100
  ) |> 
  report_rope(phase, group_2, contrast)
```


### Frequentist

```{r}
#| label: vviq-2-groups
#| eval: false

mf_nl_vviq_2 <-
  mgcv::bam(
    formula = rt ~
      group_2_category +
      s(term, by = group_2_category,  bs = "tp", k = 4) +
      s(term, problem, by = group_2,  bs = "fs", m = 1, k = 4) +
      s(term, id,      by = category, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_vviq_2 <-
  mf_nl_vviq_2 |>
  get_contrast(
    ~ group_2_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = FALSE,
    adjust = "none"
  )
```

The chunk above does not run, but the contrasts can be accessed natively in the
package in the `nl_contrasts` object. The code below tweaks them a bit for a
prettier display:

```{r}
#| label: contrasts-vviq-2

nl_contrasts$vviq_2 |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    contrast, " - ", names = c("group_cat_1", "group_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_1, ".", names = c("group_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_2, ".", names = c("group_2", "category_2")
  ) |>
  dplyr::filter(group_1 == group_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(group = group_1, `RT difference` = estimate) |>
  dplyr::arrange(term, group) |>
  dplyr::mutate(
    term = term |>
      as.character() |>
      dplyr::case_match(
        "1" ~ "Premise 1",
        "2" ~ "Premise 2",
        "3" ~ "Premise 3",
        "4" ~ "Conclusion"
      ),
  ) |>
  knitr::kable(digits = 3)
```

Let's do the same for the other two classifications.

## VVIQ 3 groups

```{r}
#| label: vviq-3-groups
#| eval: false

mf_nl_vviq_3 <-
  mgcv::bam(
    formula = rt ~
      group_3_category +
      s(term, by = group_3_category,  bs = "tp", k = 4) +
      s(term, problem, by = group_3,  bs = "fs", m = 1, k = 4) +
      s(term, id,      by = category, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_vviq_3 <-
  mf_nl_vviq_3 |>
  get_contrast(
    ~ group_3_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = FALSE,
    adjust = "none"
  )
```

Same process as above, contrasts are saved in the package:

```{r}
#| label: contrasts-vviq-3

nl_contrasts$vviq_3 |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    contrast, " - ", names = c("group_cat_1", "group_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_1, ".", names = c("group_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_2, ".", names = c("group_2", "category_2")
  ) |>
  dplyr::filter(group_1 == group_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(group = group_1, `RT difference` = estimate) |>
  dplyr::arrange(term, group) |>
  dplyr::mutate(
    term = term |>
      as.character() |>
      dplyr::case_match(
        "1" ~ "Premise 1",
        "2" ~ "Premise 2",
        "3" ~ "Premise 3",
        "4" ~ "Conclusion"
      ),
  ) |>
  knitr::kable(digits = 3)
```

## OSIVQ 3 clusters

```{r}
#| label: osivq-3-clusters
#| eval: false

mf_nl_osivq <-
  mgcv::bam(
    formula = rt ~
      cluster_category +
      s(term, by = cluster_category,    bs = "tp", k = 4) +
      s(term, problem, by = cluster,    bs = "fs", m = 1, k = 4) +
      s(term, id,      by = category, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_osivq <-
  mf_nl_osivq |>
  get_contrast(
    ~ cluster_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = TRUE
  )
```

Same process as above, contrasts are saved in the package:

```{r}
#| label: contrasts-osivq

nl_contrasts$osivq |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    cluster_category_pairwise, " - ", names = c("cluster_cat_1", "cluster_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    cluster_cat_1, ".", names = c("cluster_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    cluster_cat_2, ".", names = c("cluster_2", "category_2")
  ) |>
  dplyr::filter(cluster_1 == cluster_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(cluster = cluster_1, `RT difference` = estimate) |>
  dplyr::arrange(term, cluster) |>
  dplyr::mutate(
    term = dplyr::case_match(
      term,
      1 ~ "Premise 1",
      2 ~ "Premise 2",
      3 ~ "Premise 3",
      4 ~ "Conclusion"
    ),
  ) |>
  knitr::kable(digits = 3)
```

# Visualisation

The non-linear dynamics were represented using model predictions for each term,
grouping and category. These predictions have been computed with the
`modelbased::estimate_relation()` function and the code below (which does not
run here):

```{r}
#| label: model-predictions
#| eval: false

# Getting model predictions (for plotting)
preds_2 <-
  modelbased::estimate_relation(
    mf_nl_vviq_2,
    by = c("group_2_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(group_2_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    group_2_category,
    delim = ".",
    names = c("group", "category")
  )

preds_3 <-
  modelbased::estimate_relation(
    mf_nl_vviq_3,
    by = c("group_3_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(group_3_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    group_3_category,
    delim = ".",
    names = c("group", "category")
  )

preds_osivq <-
  modelbased::estimate_relation(
    mf_nl_osivq,
    by = c("cluster_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(cluster_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    cluster_category,
    delim = ".",
    names = c("group", "category")
  )
```

... And have been saved as package data in the `nl_predictions` object. We
created a `plot_nl()` function to visualise the data easily and added
significance labels based on the contrasts with the `add_significance()` helper
function (which is why the code below is so lengthy).

```{r}
#| label: plot-nl
#| fig.width: 9
#| fig.height: 9
#| fig.alt: >
#|  Plots showing the mean RTs in each trial term and category for several
#|  different grouping variables.
#| warning: false
#| message: false

library(patchwork)

pnl1 <-
  plot_nl(
    nl_predictions$vviq_2, title = "VVIQ 2 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )  +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 1.93,
      y_star = 5.97,
      stars  = "°",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 5.9
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 1.93,
      y_star = 6.24,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.17
    )
  ) +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.06,
      y_star = 7.87,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.8
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.01,
      y_star = 8.27,
      stars  = "°",
      x_line = .data$x_star - 0.1,
      x_line_end = .data$x_star + 0.1,
      y_line = 8.2
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 3.06,
      y_star = 8.47,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.4
    )
  )  +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  )

pnl2 <-
  plot_nl(
    nl_predictions$vviq_3,
    title = "VVIQ 3 groups",
    plot.margin = ggplot2::margin(t = 10),
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  ) +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 1.93,
      y_star = 6.63,
      stars  = "*",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.57
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 1.93,
      y_star = 6.24,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.17
    )
  )  +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.06,
      y_star = 8.32,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.25
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Hypophantasia")),
      x_star = 3.06,
      y_star = 7.62,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.55
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 3.06,
      y_star = 8.47,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.4
    )
  )  +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  )

pnl3 <-
  plot_nl(
    nl_predictions$osivq,
    title = "OSIVQ clusters",
    plot.margin = ggplot2::margin(t = 10),
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  ) +

  # 1st premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 1,
      y_star = 8.47,
      stars  = "°",
      x_line = .data$x_star - 0.1,
      x_line_end = .data$x_star + 0.1,
      y_line = 8.4
    )
  ) +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 1.93,
      y_star = 6.37,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.3
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Verbaliser")),
      x_star = 1.93,
      y_star = 6.17,
      stars  = "°",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.1
    )
  ) +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 3.06,
      y_star = 8.62,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.55
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Verbaliser")),
      x_star = 3.06,
      y_star = 8.04,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.97
    )
  )

pnl <- pnl1 / pnl2 / pnl3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(pnl)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
