---
title: "Exploring RTs per phase (non-linear modelling)"
output: rmarkdown::html_vignette
bibliography: "references.bib"
csl: "apa.csl"
vignette: >
  %\VignetteIndexEntry{Exploring RTs per phase (non-linear modelling)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the exploratory non-linear modelling
conducted to analyse the evolution of participants' response times over the
course of the trial phases. Indeed, in addition to the total response times per
trial usually analysed in VIIE studies, we collected response time data for each
trial phase (three premises and the conclusion). We explored whether the VIIE
could be specific to certain phases, or whether it was a difference in the
dynamics of reasoning across trial phases rather than an overall difference in
speed, by modelling response times throughout a trial with non-linear models.

The non-linear modelling used here was heavily inspired by [this tutorial by
Ping Hei Yeung](https://yeungpinghei.github.io/) and [this Stackoverflow
thread](https://stackoverflow.com/q/68659805). Only Bayesian analyses were
reported in the manuscript for brevity, but equivalent frequentist analyses were
also conducted to test the convergence of the two approaches on similar results.
They are reported in this vignette alongside the Bayesian analyses for
completeness.

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready and clustered data (see
`vignette("preparing_data")` and `vignette("osivq_clusters")` for details) and
filter trials based on RTs (see `vignette("analysing_rt")` for details).

```{r}
#| label: get-data

df_rt <-
  get_clustered_data("experiment") |>
  filter_trials_on_rt() |> 
  dplyr::select(id:conclusion_rt)

dplyr::glimpse(df_rt)
```

By default, the `df_rt` data frame has one variable (one column) containing the
response times of each participant for each trial phase. We needed to have all
of these phases (premises and conclusion) in a single "phase" variable along
with the associated "rt" variable for modelling, so we gathered these four
columns into two long columns by "pivoting" the RT data in a long format
(reducing the number of columns and increasing the number of rows). This
operation is performed by the `pivot_phases_longer()` function.

We also needed to create "interaction" variables between the grouping variables
and the categories of the problems (visual, spatial, control) using the
`base::interaction()` function because *mgcv* models used for non-linear
modelling do not support interaction terms natively.

```{r}
#| label: pivot-terms

df_rt_long <-
  pivot_phases_longer(df_rt) |>
  dplyr::mutate(
    group_2_category = interaction(group_2, category),
    group_3_category = interaction(group_3, category),
    cluster_category = interaction(cluster, category)
  )

dplyr::glimpse(df_rt_long)
```

# Method

We fitted generalised additive models using the *brms* package for Bayesian
models [@burknerBrmsPackageBayesian2017] or the *mgcv* package [@mgcv2011] to
predict response times with a grouping variable (VVIQ groups, OSIVQ clusters),
Category (visual, spatial or control) and their interaction as fixed categorical
predictors, as well as “smooth” (non-linear) terms that capture the evolution of
RTs across trial phases (premise 1/2/3 and conclusion) for each grouping and
category and each participant (the latter two being equivalent to a “random
intercept” in mixed models).

Let's break this down.

## Grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_rt_long` data frame:

-   `group_4` is the 4-group VVIQ classification with an "aphantasia" group
    (VVIQ = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group
    (VVIQ $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It
    was not used in the analyses because the hyperphantasia group was too small
    (N = 4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

The same modelling pipeline was therefore applied three times, once for each of
the last three grouping variables.

## Bayesian models

Bayesian models were fitted using the `brms::brm()` function through a custom
wrapper, `fit_brms_model()`, that sets several default options for us. 24000
post-warmup iterations were spread across all available CPU for parallel
processing[^1], with 2000 additional warmup iterations per chain. A fixed seed
was used for reproducibility. The [CmdStanR](https://mc-stan.org/cmdstanr/)
back-end was used for better performance. CmdStanR needs a special installation
command which is provided in the next chunk if needed. Regularising priors
coefficients were used to improve convergence and avoid overfitting. A normal
prior with mean 0 and standard deviation 1 was used on the fixed effects, an
exponential prior with rate 1 on the standard deviation of the random effects,
and an exponential prior with rate 1 on the residual standard deviation. The
fine tuning of these priors was done based on prior predictive checks (see
`vignette("osivq_clusters")` for details). To avoid having to refit the models
each time the vignette is built and improve reproducibility, fitted models
(which took more than 6 hours per model on 24 cores) are saved in the
`vignettes/models` folder and loaded in the R chunks.

[^1]: This specific number was chosen arbitrarily to have an even number of
    iterations on each chain on the two computers that were used for the
    analyses, one with 20 cores and one with 24 cores. The only important aspect
    here is to have enough iterations to ensure convergence and a good effective
    sample size.

Finally, we tested our hypotheses on group and category differences with
marginal contrasts. This task was performed with functions from the
*marginaleffects* package. We used the `marginaleffects::avg_comparisons()`
function to compute contrasts and a custom `report_rope()` wrapper (around
`marginaleffects::posterior_draws()`, `bayestestR::rope_range()` and
`bayestestR::p_direction()`) to summarise their posterior distributions. We
computed the probability of direction (PD) of the contrasts and the proportions
of their posterior distributions below, inside or above a region of practical
equivalence to the null (ROPE) [see @makowskiIndicesEffectExistence2019].

The setup of the CmdStanR back-end, marginaleffects options and the model priors
is done in the chunk below.

```{r}
#| label: bayesian-setup
#| message: false
#| warning: false

# if(!requireNamespace("cmdstanr", quietly = TRUE)) {
#   install.packages(
#     "cmdstanr",
#     repos = c('https://stan-dev.r-universe.dev', getOption("repos"))
#   )
# }
options("marginaleffects_safe" = FALSE)
draws <- seq(1, 2000, 1) # To limit draws that will be used for marginaleffects

prior_nl <- c(
  brms::prior(normal(0, 1), class = "b"),
  brms::prior(exponential(1), class = "sds"),
  brms::prior(exponential(1), class = "sigma")
)
```

## Frequentist models

The models were fitted using the `mgcv::bam()` function. After model fit, we
tested group and category differences with marginal contrasts. This task was
performed with the `get_contrast()` function, which is a wrapper around several
functions from the *emmeans* package.

These models took a long time to fit and needed to be saved for ease of reuse,
but model fits were too heavy to be kept as package data (more than 100 MB).
Thus, two sub-products of the models were saved as package data instead:

-   Contrast analyses obtained with `get_contrast()` for the statistical tests.
    The analyses for the three grouping variables have been saved in the
    `nl_contrasts` list.

-   Model predictions obtained with `modelbased::estimate_relation()` for the
    data used for visualisation. The predictions for the three grouping
    variables have been saved in the `nl_predictions` list.

The code to fit the frequentist models and get the contrasts is provided below
but does not run.

# Results

## VVIQ 2 groups

### Bayesian

```{r}
#| label: bayesian-vviq-2-groups
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_nl_vviq_2 <-
  fit_brms_model(
    formula = rt ~ 
      group_2 * category + 
      s(phase, by = interaction(group_2, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_vviq_2.rds"
  )

# Posterior predictive check (best model performance indicator)
mb_nl_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)

# Contrasts
marginaleffects::avg_comparisons(
  mb_nl_vviq_2, 
  variables = list("category" = "pairwise"), 
  by = c("group_2", "phase"), 
  draw_ids = draws
  ) |> 
  report_rope(phase, group_2, contrast) |> knitr::kable()
```

### Frequentist

Note that the frequentist models were fitted with an earlier version of the
package when the `term` variable was used instead of `phase`. The code below
reflects this older version for consistency with the saved contrasts. Replacing
`term` with `phase` would not change the results (but would be required to refit
the models).

```{r}
#| label: frequentist-vviq-2-groups
#| eval: false

mf_nl_vviq_2 <-
  mgcv::bam(
    formula = rt ~
      group_2_category +
      s(term, by = group_2_category,  bs = "tp", k = 4) +
      s(term, id, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_vviq_2 <-
  mf_nl_vviq_2 |>
  get_contrast(
    ~ group_2_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = FALSE,
    adjust = "none"
  )
```

The chunk above does not run, but the contrasts can be accessed natively in the
package in the `nl_contrasts` object. The code below tweaks them a bit for a
prettier display:

```{r}
#| label: frequentist-contrasts-vviq-2
#| message: false
#| warning: false

nl_contrasts$vviq_2 |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    contrast, " - ", names = c("group_cat_1", "group_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_1, ".", names = c("group_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_2, ".", names = c("group_2", "category_2")
  ) |>
  dplyr::filter(group_1 == group_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(group = group_1, `RT difference` = estimate) |>
  dplyr::arrange(term, group) |>
  dplyr::mutate(
    term = term |>
      as.character() |>
      dplyr::case_match(
        "1" ~ "Premise 1",
        "2" ~ "Premise 2",
        "3" ~ "Premise 3",
        "4" ~ "Conclusion"
      ),
  ) |>
  knitr::kable(digits = 3)
```

Let's do the same for the other two classifications.

## VVIQ 3 groups

### Bayesian

```{r}
#| label: bayesian-vviq-3-groups
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_nl_vviq_3 <-
  fit_brms_model(
    formula = rt ~ 
      group_3 * category + 
      s(phase, by = interaction(group_3, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_vviq_3.rds"
  )

# Posterior predictive check (best model performance indicator)
mb_nl_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)

# Contrasts
marginaleffects::avg_comparisons(
  mb_nl_vviq_3, 
  variables = list("category" = "pairwise"), 
  by = c("group_3", "phase"), 
  draw_ids = draws
  ) |> 
  report_rope(phase, group_3, contrast) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-3-groups
#| eval: false

mf_nl_vviq_3 <-
  mgcv::bam(
    formula = rt ~
      group_3_category +
      s(term, by = group_3_category,  bs = "tp", k = 4) +
      s(term, id, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_vviq_3 <-
  mf_nl_vviq_3 |>
  get_contrast(
    ~ group_3_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = FALSE,
    adjust = "none"
  )
```

Same process as above, contrasts are saved in the package:

```{r}
#| label: frequentist-contrasts-vviq-3
#| message: false
#| warning: false

nl_contrasts$vviq_3 |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    contrast, " - ", names = c("group_cat_1", "group_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_1, ".", names = c("group_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    group_cat_2, ".", names = c("group_2", "category_2")
  ) |>
  dplyr::filter(group_1 == group_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(group = group_1, `RT difference` = estimate) |>
  dplyr::arrange(term, group) |>
  dplyr::mutate(
    term = term |>
      as.character() |>
      dplyr::case_match(
        "1" ~ "Premise 1",
        "2" ~ "Premise 2",
        "3" ~ "Premise 3",
        "4" ~ "Conclusion"
      ),
  ) |>
  knitr::kable(digits = 3)
```

## OSIVQ 3 clusters

### Bayesian

```{r}
#| label: bayesian-osivq-3-clusters
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_nl_osivq <-
  fit_brms_model(
    formula = rt ~ 
      cluster * category + 
      s(phase, by = interaction(cluster, category), k = 4) +
      s(phase, id, bs = "fs", m = 1, k = 4),
    data    = df_rt_long,
    family  = brms::shifted_lognormal(),
    prior   = prior_nl,
    refresh = 10,
    file    = "models/m_nl_osivq.rds"
  )

# Posterior predictive check (best model performance indicator)
mb_nl_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)

# Contrasts
marginaleffects::avg_comparisons(
  mb_nl_osivq, 
  variables = list("category" = "pairwise"), 
  by = c("cluster", "phase"), 
  draw_ids = draws
  ) |> 
  report_rope(phase, cluster, contrast) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-osivq-3-clusters
#| eval: false

mf_nl_osivq <-
  mgcv::bam(
    formula = rt ~
      cluster_category +
      s(term, by = cluster_category,    bs = "tp", k = 4) +
      s(term, id, bs = "fs", m = 1, k = 4),
    family  = Gamma(link = "identity"),
    data    = df_rt_long,
    method  = "fREML"
  )

contrasts_osivq <-
  mf_nl_osivq |>
  get_contrast(
    ~ cluster_category | term,
    at = list(term = c(1, 2, 3, 4)),
    interaction = TRUE
  )
```

Same process as above, contrasts are saved in the package:

```{r}
#| label: frequentist-contrasts-osivq
#| message: false
#| warning: false

nl_contrasts$osivq |>
  as.data.frame() |>
  tidyr::separate_wider_delim(
    cluster_category_pairwise, " - ", names = c("cluster_cat_1", "cluster_cat_2")
  ) |>
  tidyr::separate_wider_delim(
    cluster_cat_1, ".", names = c("cluster_1", "category_1")
  ) |>
  tidyr::separate_wider_delim(
    cluster_cat_2, ".", names = c("cluster_2", "category_2")
  ) |>
  dplyr::filter(cluster_1 == cluster_2 & p.value < 1) |>
  dplyr::select(!c(
    tidyselect::contains("group_2"),
    tidyselect::contains("cluster_2"),
    tidyselect::contains("SE"),
    tidyselect::contains("df")
  )) |>
  dplyr::mutate(across(c(estimate:p.value), ~round(., 3))) |>
  tidyr::unite(`Category contrast`, category_1, category_2, sep = " - ") |>
  dplyr::rename(cluster = cluster_1, `RT difference` = estimate) |>
  dplyr::arrange(term, cluster) |>
  dplyr::mutate(
    term = dplyr::case_match(
      term,
      1 ~ "Premise 1",
      2 ~ "Premise 2",
      3 ~ "Premise 3",
      4 ~ "Conclusion"
    ),
  ) |>
  knitr::kable(digits = 3)
```

# Visualisation

The non-linear dynamics were represented using model predictions for each phase,
grouping and category. These predictions have been computed with the
`modelbased::estimate_relation()` function on the frequentist models and the
code below (which does not run here). The frequentist models are used here for
convenience, but the Bayesian models could also be used to get similar
predictions.

```{r}
#| label: model-predictions
#| eval: false

# Getting model predictions (for plotting)
preds_2 <-
  modelbased::estimate_relation(
    mf_nl_vviq_2,
    by = c("group_2_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(group_2_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    group_2_category,
    delim = ".",
    names = c("group", "category")
  )

preds_3 <-
  modelbased::estimate_relation(
    mf_nl_vviq_3,
    by = c("group_3_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(group_3_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    group_3_category,
    delim = ".",
    names = c("group", "category")
  )

preds_osivq <-
  modelbased::estimate_relation(
    mf_nl_osivq,
    by = c("cluster_category", "term")
  ) |>
  as.data.frame() |>
  dplyr::select(cluster_category, term, Predicted, CI_low, CI_high) |>
  tidyr::separate_wider_delim(
    cluster_category,
    delim = ".",
    names = c("group", "category")
  )
```

They have been saved as package data in the `nl_predictions` object. We created
a `plot_nl()` function to visualise the data easily and added significance
labels based on the contrasts with the `add_significance()` helper function
(which is why the code below is so lengthy).

```{r}
#| label: plot-nl
#| fig.width: 9
#| fig.height: 9
#| fig.alt: >
#|  Plots showing the mean RTs in each trial term and category for several
#|  different grouping variables.
#| warning: false
#| message: false

library(patchwork)

pnl1 <-
  plot_nl(
    nl_predictions$vviq_2, title = "VVIQ 2 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )  +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 1.93,
      y_star = 5.97,
      stars  = "°",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 5.9
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 1.93,
      y_star = 6.24,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.17
    )
  ) +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.06,
      y_star = 7.87,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.8
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.01,
      y_star = 8.27,
      stars  = "°",
      x_line = .data$x_star - 0.1,
      x_line_end = .data$x_star + 0.1,
      y_line = 8.2
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 3.06,
      y_star = 8.47,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.4
    )
  )  +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  )

pnl2 <-
  plot_nl(
    nl_predictions$vviq_3,
    title = "VVIQ 3 groups",
    plot.margin = ggplot2::margin(t = 10),
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  ) +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 1.93,
      y_star = 6.63,
      stars  = "*",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.57
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 1.93,
      y_star = 6.24,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.17
    )
  )  +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Aphantasia")),
      x_star = 3.06,
      y_star = 8.32,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.25
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Hypophantasia")),
      x_star = 3.06,
      y_star = 7.62,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.55
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 3.06,
      y_star = 8.47,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.4
    )
  )  +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Typical")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  )

pnl3 <-
  plot_nl(
    nl_predictions$osivq,
    title = "OSIVQ clusters",
    plot.margin = ggplot2::margin(t = 10),
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  ) +

  # 1st premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 1,
      y_star = 8.47,
      stars  = "°",
      x_line = .data$x_star - 0.1,
      x_line_end = .data$x_star + 0.1,
      y_line = 8.4
    )
  ) +

  # 2nd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 1.93,
      y_star = 6.37,
      stars  = "**",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.3
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Verbaliser")),
      x_star = 1.93,
      y_star = 6.17,
      stars  = "°",
      x_line = .data$x_star - 0.06,
      x_line_end = .data$x_star + 0.06,
      y_line = 6.1
    )
  ) +

  # 3rd premise
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 3.06,
      y_star = 8.62,
      stars  = "***",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 8.55
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Visualiser")),
      x_star = 2.93,
      y_star = 7.77,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.7
    )
  ) +
  add_significance(
    size_star = 3,
    tibble::tibble(
      group  = factor(c("Verbaliser")),
      x_star = 3.06,
      y_star = 8.04,
      stars  = "*",
      x_line = .data$x_star - 0.05,
      x_line_end = .data$x_star + 0.05,
      y_line = 7.97
    )
  )

pnl <- pnl1 / pnl2 / pnl3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(pnl)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
