---
title: "Accuracy analyses"
output: rmarkdown::html_vignette
message: false
warning: false
bibliography: "references.bib"
csl: "apa.csl"
vignette: >
  %\VignetteIndexEntry{Accuracy analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the analyses of participants'
accuracy on the reasoning problems. Only Bayesian analyses were reported in the
manuscript for brevity, but equivalent frequentist analyses were also conducted
to test the convergence of the two approaches on similar results. They are
reported in this vignette alongside the Bayesian analyses for completeness.

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready and clustered data (see
`vignette("preparing_data")` and `vignette("osivq_clusters")` for details).

```{r}
#| label: get-data

df_expe <-
  get_clustered_data("experiment") |> 
  dplyr::select(id, group_4:strategy_group, problem, category, accuracy)

dplyr::glimpse(df_expe)
```

# Method

We fitted hierarchical linear models (also said "multilevel" or "mixed" models)
with binomial distributions and logit links using the *brms* package for
Bayesian models [@burknerBrmsPackageBayesian2017] or the *glmmTMB* package for
frequentist models [@glmmTMB2025]. The models included a grouping variable (VVIQ
groups, OSIVQ clusters), Category (visual, spatial, or control) along with their
two-way interactions as fixed categorical predictors. Varying slopes and
intercepts (“random effects”) have been added for each participant by category
along with an intercept for each problem.

Let's break this down.

## Grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_expe` data frame:

-   `group_4` is the 4-group VVIQ classification with an "aphantasia" group
    (VVIQ = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group
    (VVIQ $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It
    was not used in the analyses because the hyperphantasia group was too small
    (N = 4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

-   `strategy_group` is a classification based on the self-reported strategies
    used by the participants to solve the problems. It focuses on whether
    participants used a visual mental imagery strategy and contains two groups:
    "Visual strategy user" and "No visual strategy". It is not reported in the
    manuscript because it was suggested to us after submission by colleagues at
    a conference.

The same modelling pipeline was therefore applied four times, once for each of
the last four grouping variables.

## Model formula

We created a small `build_formula()` helper function to write the formula
easily, as we used the same model structure a lot of times:

```{r}
#| label: build-formula

build_formula("accuracy", "group_2")

build_formula("accuracy", "cluster")
```

## Bayesian models

Bayesian models were fitted using the `brms::brm()` function through a custom
wrapper, `fit_brms_model()`, that sets several default options for us. 24000
post-warmup iterations were spread across all available CPU for parallel
processing[^1], with 2000 additional warmup iterations per chain. A fixed seed
was used for reproducibility. The [CmdStanR](https://mc-stan.org/cmdstanr/)
back-end was used for better performance. CmdStanR needs a special installation
command which is provided in the next chunk if needed. A regularising normal
prior with mean 0 and standard deviation 1 was used on the fixed effects
coefficients to improve convergence and avoid overfitting. To avoid having to
refit the models each time the vignette is built and improve reproducibility,
fitted models are saved in the `vignettes/models` folder and loaded in the R
chunks.

[^1]: This specific number was chosen arbitrarily to have an even number of
    iterations on each chain on the two computers that were used for the
    analyses, one with 20 cores and one with 24 cores. The only important aspect
    here is to have enough iterations to ensure convergence and a good effective
    sample size.

After model fit, we checked for potential singularity issues and model
performance using the `get_singularity()` and `get_performance()` functions
created for the occasion (which are convenient wrappers around the *performance*
package). Posterior predictive checks were also performed with the
`performance::check_predictions()` function to check that the models were able
to reproduce the data well.

Finally, we tested our hypotheses with marginal contrasts. This task was
performed with functions from the *marginaleffects* package. We used the
`marginaleffects::avg_comparisons()` function to compute contrasts and a custom
`report_rope()` wrapper (around `marginaleffects::posterior_draws()`,
`bayestestR::rope_range()` and `bayestestR::p_direction()`) to summarise their
posterior distributions. We computed the probability of direction (PD) of the
contrasts and the proportions of their posterior distributions below, inside or
above a region of practical equivalence to the null (ROPE) [see
@makowskiIndicesEffectExistence2019]. We computed accuracy contrasts between
groups, contrasts between categories within each group, and interaction
contrasts (differences in category contrasts between groups).

The setup of the CmdStanR back-end, marginaleffects options and the accuracy
models prior is done in the chunk below.

```{r}
#| label: bayesian-setup
#| message: false
#| warning: false

if(!requireNamespace("cmdstanr", quietly = TRUE)) {
  install.packages(
    "cmdstanr",
    repos = c('https://stan-dev.r-universe.dev', getOption("repos"))
  )
}
options("marginaleffects_safe" = FALSE)
draws <- seq(1, 100, 1) # To limit draws that will be used for marginaleffects

prior_acc <- c(brms::prior(normal(0, 1), class = "b"))
```

## Frequentist models

Frequentist models were fitted using the `glmmTMB::glmmTMB()` function (and a
little `set_ranef_prior()` helper function to set a weakly informative prior on
the random effects, which helps with convergence and singularity issues). After
model fit, we checked for potential singularity issues and model performance
using the `get_singularity()` and `get_performance()` functions.

Finally, we tested our hypotheses with marginal contrasts. This task was
performed with the `report_contrast()` function, which is a wrapper around
several functions from the *emmeans* package[^2]. We computed accuracy contrasts
between groups, contrasts between categories within each group, and interaction
contrasts (differences in category contrasts between groups).

[^2]: The marginaleffects package was not used for frequentist analyses because
    it conflicted with glmmTMB models, whereas the emmeans package worked fine.

Here we go!

# Results

## VVIQ 2 groups

### Bayesian

```{r}
#| label: bayesian-vviq-2-groups
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_acc_vviq_2 <-
  fit_brms_model(
    formula = build_formula("accuracy", "group_2"),
    data    = df_expe,
    family  = brms::bernoulli(),
    prior   = prior_acc,
    file    = "models/m_acc_vviq_2.rds"
  )

# Singularity check
mb_acc_vviq_2 |> get_singularity()
# Model performance indices
# mb_acc_vviq_2  |> 
#   get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
#   knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_acc_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_acc_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_2" = "pairwise"), 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_acc_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"), by = "group_2", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(group_2, contrast) |> knitr::kable()
# Interaction contrasts
mb_acc_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "group_2",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-2-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_acc_vviq_2 <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "group_2"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(55)
  )

# Singularity
mf_acc_vviq_2 |> get_singularity()

# Performance
mf_acc_vviq_2 |> get_performance() |> knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_acc_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mf_acc_vviq_2 |> report_contrast(~ group_2, ratios = FALSE) |> knitr::kable()
 # Category contrasts within groups
mf_acc_vviq_2 |> 
  report_contrast(~ category | group_2, ratios = FALSE) |> knitr::kable()
# Interaction contrasts
mf_acc_vviq_2 |> 
  report_contrast(~ category * group_2, interaction = TRUE) |>
  knitr::kable()
```

## VVIQ 3 groups

### Bayesian

```{r}
#| label: bayesian-vviq-3-groups
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_acc_vviq_3 <-
  fit_brms_model(
    formula = build_formula("accuracy", "group_3"),
    data    = df_expe,
    family  = brms::bernoulli(),
    prior   = prior_acc,
    file    = "models/m_acc_vviq_3.rds"
  )

# Singularity check
mb_acc_vviq_3 |> get_singularity()
# Model performance indices
# mb_acc_vviq_3  |> 
#   get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
#   knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_acc_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_acc_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_3" = "pairwise"), 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_acc_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"), by = "group_3", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(group_3, contrast) |> knitr::kable()
# Interaction contrasts
mb_acc_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "group_3",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-3-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_acc_vviq_3 <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "group_3"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(20)
  )

mf_acc_vviq_3 |> get_singularity()

mf_acc_vviq_3 |> get_performance() |> knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_acc_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_acc_vviq_3 |> report_contrast(~ group_3, ratios = FALSE) |> knitr::kable()
mf_acc_vviq_3 |> report_contrast(~ category | group_3, ratios = FALSE) |> knitr::kable()
mf_acc_vviq_3 |> report_contrast(~ category * group_3, interaction = TRUE) |>
  knitr::kable()
```

## OSIVQ 3 clusters

### Bayesian

```{r}
#| label: bayesian-osivq-3-clusters
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_acc_osivq <-
  fit_brms_model(
    formula = build_formula("accuracy", "cluster"),
    data    = df_expe,
    family  = brms::bernoulli(),
    prior   = prior_acc,
    file    = "models/m_acc_osivq.rds"
  )

# Singularity check
mb_acc_osivq |> get_singularity()
# Model performance indices
# mb_acc_osivq  |> 
#   get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
#   knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_acc_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_acc_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("cluster" = "pairwise"), 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_acc_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"), by = "cluster", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(cluster, contrast) |> knitr::kable()
# Interaction contrasts
mb_acc_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "cluster",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-osivq-3-clusters
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_acc_osivq <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "cluster"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(15)
  )

mf_acc_osivq  |> get_singularity()

mf_acc_osivq  |> get_performance() |> knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_acc_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_acc_osivq  |> report_contrast(~ cluster, ratios = FALSE) |> knitr::kable()
mf_acc_osivq  |> report_contrast(~ category | cluster, ratios = FALSE) |> knitr::kable()
mf_acc_osivq  |> report_contrast(~ category * cluster, interaction = TRUE) |>
  knitr::kable()
```

## Strategy groups

### Bayesian

```{r}
#| label: bayesian-strategy-groups
#| message: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_acc_strat <-
  fit_brms_model(
    formula = build_formula("accuracy", "strategy_group"),
    data    = df_expe,
    family  = brms::bernoulli(),
    prior   = prior_acc,
    file    = "models/m_acc_strat.rds"
  )

# Singularity check
mb_acc_strat |> get_singularity()
# Model performance indices
# mb_acc_strat  |> 
#   get_performance(metrics = c("WAIC", "R2", "RMSE")) |> 
#   knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_acc_strat |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts
mb_acc_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy_group" = "pairwise"), 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(contrast) |> knitr::kable()
# Category contrasts within groups
mb_acc_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"), by = "strategy_group", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(strategy_group, contrast) |> knitr::kable()
# Interaction contrasts
mb_acc_strat |> 
  marginaleffects::avg_comparisons(
    variables = list("category" = "pairwise"),
    by = "strategy_group",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-strategy-groups
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mf_acc_strat <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "strategy_group"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(55)
  )

mf_acc_strat |> get_singularity()

mf_acc_strat |> get_performance() |> knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mf_acc_strat |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
mf_acc_strat |> report_contrast(~ strategy_group, ratios = FALSE) |> knitr::kable()
mf_acc_strat |> report_contrast(~ category | strategy_group, ratios = FALSE) |> knitr::kable()
mf_acc_strat |> report_contrast(~ category * strategy_group, interaction = TRUE) |>
  knitr::kable()
```

# Visualisation

The figures showing the distribution of accuracy data displayed in the
manuscript were created with the `plot_superb_jitter()` function from the
package, which wraps functions from the *ggplot2* and *superb* packages to
create nice visualisations easily. A little `add_significance()` helper function
was also created to add significance stars to the plots.

```{r}
#| label: plot-accuracy
#| fig.width: 10
#| fig.height: 4
#| fig.alt: >
#|  A jitter plot showing the mean accuracy in each category for several
#|  different grouping variables.

library(patchwork)
library(superb)

pa1 <-
  plot_superb_jitter(
    df_expe, accuracy, group_2,
    title = "VVIQ 2 groups", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 2,
      y_star = 1.07,
      stars  = "*",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa2 <-
  plot_superb_jitter(
    df_expe, accuracy, group_3,
    title = "VVIQ 3 groups", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 1.07,
      stars  = "*",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa3 <-
  plot_superb_jitter(
    df_expe, accuracy, cluster,
    title = "OSIVQ clusters", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 1.065,
      stars  = "°",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa <- pa1 + pa2 + pa3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(pa)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
