---
title: "Strategy use analyses"
output: rmarkdown::html_vignette
message: false
warning: false
bibliography: "references.bib"
csl: "apa.csl"
vignette: >
  %\VignetteIndexEntry{Strategy use analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the analyses of the self-reported
(using Likert scales) mental strategies that participants used to solve the
reasoning problems. Only Bayesian analyses were reported in the manuscript for
brevity, but equivalent frequentist analyses were also conducted to test the
convergence of the two approaches on similar results. They are reported in this
vignette alongside the Bayesian analyses for completeness.

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready and clustered data (see
`vignette("preparing_data")` and `vignette("osivq_clusters")` for details).

```{r}
#| label: get-data

df_survey <- get_clustered_data("survey") 
```

By default, the `df_survey` data frame has one variable (one column) containing
the responses of each participant for each strategy question (Visual, Spatial,
Verbal, Semantic, Sensorimotor). We needed to have all of these five modalities
in a single "strategy" variable along with the associated "score" variable for
modelling, so we gathered these five columns into two long columns by "pivoting"
the strategies data in a long format (reducing the number of columns and
increasing the number of rows). This operation is performed by the
`pivot_strategies_longer()` function.

```{r}
#| label: pivot-strategies

df_strats_long <- pivot_strategies_longer(df_survey, ordered = TRUE)

dplyr::glimpse(df_strats_long)
```

# Method

Ordinal cumulative link regression models were fitted, using the *brms* package
for Bayesian models [@burknerBrmsPackageBayesian2017] or the *ordinal* package
[@ordinal2023] for frequentist models, to predict the score (on a question about
the use of a given strategy) with a grouping variable (VVIQ groups, OSIVQ
clusters), Strategy (visual, verbal, spatial, semantic or sensorimotor) and
their two-way interaction as fixed categorical predictors. We planned to analyse
the contrasts between groups for each strategy separately.

Let's break this down.

## Grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_strats_long` data frame:

-   `group_4` is the 4-group VVIQ classification with an "aphantasia" group
    (VVIQ = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group
    (VVIQ $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It
    was not used in the analyses because the hyperphantasia group was too small
    (N = 4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

The same modelling pipeline was therefore applied three times, once for each of
the last three grouping variables.

## Bayesian models

Bayesian models were fitted using the `brms::brm()` function through a custom
wrapper, `fit_brms_model()`, that sets several default options for us. 24000
post-warmup iterations were spread across all available CPU for parallel
processing[^1], with 2000 additional warmup iterations per chain. A fixed seed
was used for reproducibility. The [CmdStanR](https://mc-stan.org/cmdstanr/)
back-end was used for better performance. CmdStanR needs a special installation
command which is provided in the next chunk if needed. Since the models were
very simple and easy to fit, default priors were used. To avoid having to refit
the models each time the vignette is built and improve reproducibility, fitted
models are saved in the `vignettes/models` folder and loaded in the R chunks.

[^1]: This specific number was chosen arbitrarily to have an even number of
    iterations on each chain on the two computers that were used for the
    analyses, one with 20 cores and one with 24 cores. The only important aspect
    here is to have enough iterations to ensure convergence and a good effective
    sample size.

After model fit, we checked for potential singularity issues and model
performance using the `get_singularity()` and `get_performance()` functions
created for the occasion (which are convenient wrappers around the *performance*
package). Posterior predictive checks were also performed with the
`performance::check_predictions()` function to check that the models were able
to reproduce the data well.

Finally, we tested our hypotheses with marginal contrasts. This task was
performed with functions from the *marginaleffects* package. We used the
`marginaleffects::avg_comparisons()` function to compute contrasts and a custom
`report_rope()` wrapper (around `marginaleffects::posterior_draws()`,
`bayestestR::rope_range()` and `bayestestR::p_direction()`) to summarise their
posterior distributions. We computed the probability of direction (PD) of the
contrasts and the proportions of their posterior distributions below, inside or
above a region of practical equivalence to the null (ROPE) [see
@makowskiIndicesEffectExistence2019]. We computed strategy use contrasts between
groups for each strategy, contrasts between strategies within each group, and
interaction contrasts (differences in strategy contrasts between groups).

The setup of the CmdStanR back-end and marginaleffects options is done in the
chunk below.

```{r}
#| label: bayesian-setup
#| message: false
#| warning: false

# if(!requireNamespace("cmdstanr", quietly = TRUE)) {
#   install.packages(
#     "cmdstanr",
#     repos = c('https://stan-dev.r-universe.dev', getOption("repos"))
#   )
# }
options("marginaleffects_safe" = FALSE)
draws <- seq(1, 12000, 1) # To limit draws that will be used for marginaleffects
```

## Frequentist models

Frequentist ordinal models were fitted using the `fit_clm()` function, which is
a wrapper around the `ordinal::clm()` function that simply sets a probit link by
default. After model fit, we checked for potential singularity issues and model
performance using the `get_singularity()` and `get_performance()` functions
created for the occasion (which are convenient wrappers around the *performance*
package).

Finally, we tested our hypotheses with marginal contrasts. This task was
performed with the `report_contrast()` function, which is a wrapper around
several functions from the *emmeans* package[^2]. We used it to compute strategy
use contrasts between groups for each strategy, contrasts between strategies
within each group, and interaction contrasts (differences in strategy contrasts
between groups).

[^2]: The marginaleffects package was not used for frequentist analyses because
    it conflicted with *ordinal* models, whereas the emmeans package worked
    fine.

Here we go!

# Results

## VVIQ 2 groups

### Bayesian

```{r}
#| label: bayesian-vviq-2-groups
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_strats_vviq_2 <-
  fit_brms_model(
    formula = score ~ group_2 * strategy,
    data    = df_strats_long,
    family  = brms::cumulative(link = "probit"),
    file    = "models/m_strat_vviq_2.rds"
  )

# Singularity check
mb_strats_vviq_2 |> get_singularity()
# Model performance indices
mb_strats_vviq_2  |>
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |>
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_strats_vviq_2 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts by strategy
mb_strats_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_2" = "pairwise"), by = "strategy", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(strategy, contrast) |> knitr::kable()
# Strategy contrasts within groups
mb_strats_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"), by = "group_2", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(group_2, contrast) |> knitr::kable()
# Interaction contrasts
mb_strats_vviq_2 |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"),
    by = "group_2",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-2-groups
#| message: false
#| warning: false

mf_strats_vviq_2 <- fit_clm(score ~ group_2 * strategy, df_strats_long)

# Singularity
mf_strats_vviq_2 |> get_singularity()

# Performance
mf_strats_vviq_2 |> get_performance() |> knitr::kable(align = "c")
# Group contrasts by strategy
mf_strats_vviq_2 |> report_contrast(~ group_2 | strategy) |> knitr::kable()
# Strategy contrasts within groups
mf_strats_vviq_2 |> report_contrast(~ strategy | group_2) |> knitr::kable()
# Interaction contrasts
mf_strats_vviq_2 |> report_contrast(~ group_2 * strategy, interaction = TRUE) |>
  knitr::kable()
```

## VVIQ 3 groups

### Bayesian

```{r}
#| label: bayesian-vviq-3-groups
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_strats_vviq_3 <-
  fit_brms_model(
    formula = score ~ group_3 * strategy,
    data    = df_strats_long,
    family  = brms::cumulative(link = "probit"),
    file    = "models/m_strat_vviq_3.rds"
  )

# Singularity check
mb_strats_vviq_3 |> get_singularity()
# Model performance indices
mb_strats_vviq_3  |>
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |>
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_strats_vviq_3 |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts by strategy
mb_strats_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("group_3" = "pairwise"), by = "strategy", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(strategy, contrast) |> knitr::kable()
# Strategy contrasts within groups
mb_strats_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"), by = "group_3", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(group_3, contrast) |> knitr::kable()
# Interaction contrasts
mb_strats_vviq_3 |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"),
    by = "group_3",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-vviq-3-groups
#| message: false
#| warning: false

mf_strats_vviq_3 <- fit_clm(score ~ group_3 * strategy, df_strats_long)

mf_strats_vviq_3 |> get_singularity()

mf_strats_vviq_3 |> get_performance() |> knitr::kable(align = "c")
mf_strats_vviq_3 |> report_contrast(~ group_3 | strategy) |> knitr::kable()
mf_strats_vviq_3 |> report_contrast(~ strategy | group_3) |> knitr::kable()
mf_strats_vviq_3 |> report_contrast(~ group_3 * strategy, interaction = TRUE) |>
  knitr::kable()
```

## OSIVQ 3 clusters

### Bayesian

```{r}
#| label: bayesian-osivq-3-clusters
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 3.5
#| fig-alt: >
#|  Posterior predictive distributions.

mb_strats_osivq <-
  fit_brms_model(
    formula = score ~ cluster * strategy,
    data    = df_strats_long,
    family  = brms::cumulative(link = "probit"),
    file    = "models/m_strat_osivq.rds"
  )

# Singularity check
mb_strats_osivq |> get_singularity()
# Model performance indices
mb_strats_osivq  |>
  get_performance(metrics = c("WAIC", "R2", "RMSE")) |>
  knitr::kable(align = "c")
# Posterior predictive check (best model performance indicator)
mb_strats_osivq |> 
  performance::check_predictions() |> plot() + theme_pdf(base_size = 12)
# Group contrasts by strategy
mb_strats_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("cluster" = "pairwise"), by = "strategy", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(strategy, contrast) |> knitr::kable()
# Strategy contrasts within groups
mb_strats_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"), by = "cluster", 
    type = "link",
    draw_ids = draws
  ) |>
  report_rope(cluster, contrast) |> knitr::kable()
# Interaction contrasts
mb_strats_osivq |> 
  marginaleffects::avg_comparisons(
    variables = list("strategy" = "pairwise"),
    by = "cluster",
    hypothesis = ~revpairwise, # for the interaction
    type = "link",
    draw_ids = draws
  ) |> 
  report_rope(hypothesis) |> knitr::kable()
```

### Frequentist

```{r}
#| label: frequentist-osivq-3-clusters
#| message: false
#| warning: false

mf_strats_osivq  <- fit_clm(score ~ cluster * strategy, df_strats_long)

mf_strats_osivq |> get_singularity()

mf_strats_osivq |> get_performance() |> knitr::kable(align = "c")
mf_strats_osivq |> report_contrast(~ cluster | strategy) |> knitr::kable()
mf_strats_osivq |> report_contrast(~ strategy | cluster) |> knitr::kable()
mf_strats_osivq |> report_contrast(~ cluster * strategy, interaction = TRUE) |>
  knitr::kable()
```

# Visualisation

Strategy data were visualised using two different representations: as average
Likert scores and as proportions of answers in each Likert category for each
strategy and group. These figures were created with the functions
`plot_strategies_scores()` (for the average scores) and
`plot_strategies_barplot()` (for the proportions) from the package. A little
`add_significance()` helper function was also created to add significance stars
to the average score plots.

```{r}
#| label: plot-strategies-average
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 4
#| fig.alt: >
#|  A plot showing the mean Likert score in each strategy for several
#|  different grouping variables.


library(patchwork)

star_size <- 3

ps1 <-
  plot_strategies_scores(
    df_strats_long,
    grouping = group_2,
    title = "VVIQ 2 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )
ps2 <-
  plot_strategies_scores(
    df_strats_long,
    grouping = group_3,
    title = "VVIQ 3 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )
ps3 <-
  plot_strategies_scores(
    df_strats_long,
    grouping = cluster,
    title = "OSIVQ clusters",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )

ps <-
  ps1 + ps2 + ps3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(ps)
```

```{r}
#| label: plot-strategies-proportions
#| message: false
#| warning: false
#| fig.width: 10
#| fig.height: 8
#| fig.alt: >
#|  A plot showing the proportion of answers on each Likert score for each
#|  strategy and several different grouping variables.

pb1 <- 
  plot_strategies_barplot(
    df_strats_long, group_2, title = "VVIQ 2 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )
pb2 <- 
  plot_strategies_barplot(
    df_strats_long, group_3, title = "VVIQ 3 groups",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )
pb3 <- 
  plot_strategies_barplot(
    df_strats_long, cluster, title = "OSIVQ clusters",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA)
  )

pb <-
  pb1 + pb2 + pb3 +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(legend.position = "bottom")
  
plot(pb)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
