---
title: "Accuracy analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accuracy analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette contains a full breakdown of the analyses of participants'
accuracy on the reasoning problems. The most interesting results are
synthetically presented in the main manuscript (preprint
[here](https://doi.org/10.31234/osf.io/vsjtb_v1)). The present document aims to
provide a full account of the steps taken to reach these results, and to provide
additional analyses that were not included in the manuscript for brevity.

```{r setup}
library(aphantasiaReasoningViie)
```

# Data preparation

First, let's get the clean, analysis-ready data, create cognitive style clusters
using OSIVQ scores (see `vignette("preparing_data")` and
`vignette("osivq_clusters")` for details), and add this classification to the
dataset containing accuracy data.

```{r}
#| label: get-data

df_survey  <- get_clean_data()$df_survey
# Clustering OSIVQ data
clustering <- cluster_osivq(df_survey)
# Adding named clusters to the survey data
df_survey <- add_named_clusters(
  df_survey, clustering,
  names  = c("Spatialiser", "Visualiser", "Verbaliser"),
  levels = c("Visualiser", "Spatialiser", "Verbaliser"),
  contrasts = c("_visualiser", "_spatialiser", "_verbaliser"),
  base = 1
)
# Merging with experiment data
df_expe <-
  dplyr::left_join(
    get_clean_data()$df_expe,
    df_survey |> dplyr::select(id, cluster),
    by = dplyr::join_by("id")
  ) |>
  dplyr::relocate(cluster, .after = "group") |> 
  dplyr::select(id, group:strategy_group, problem, category, accuracy)

dplyr::glimpse(df_expe)
```

# Method

As described in the manuscript:

> We fitted Generalised Linear Mixed Models (GLMMs) with binomial distributions
> and logit links using the glmmTMB package (McGillycuddy et al., 2025) to
> predict accuracy with a grouping variable (VVIQ groups, OSIVQ clusters),
> Category (visual, spatial, or control) along with their two-way interactions
> as fixed categorical predictors. Varying slopes and intercepts (“random
> effects”) have been added for each participant by category and for each
> problem by grouping variable.

Let's break this down.

## The grouping variables

We used several grouping variables to classify participants, all of which are in
the `df_expe` data frame:

-   `group` is the 4-group VVIQ classification with an "aphantasia" group (VVIQ
    = 16), "hypophantasia" group (VVIQ $\in$ [17, 32]), "typical" group (VVIQ
    $\in$ [33, 74]), and "hyperphantasia" group (VVIQ $\in$ [75, 80]). It was
    not used in the analyses because the hyperphantasia group was too small (N =
    4).

-   `group_3` is the same as `group`, but with the hyperphantasia group merged
    with the typical group. It is referred to as "VVIQ 3 groups" in the
    manuscript results.

-   `group_2` is the binary classification often used in aphantasia literature
    that additionally merges the hypophantasia group with the aphantasia group.
    It is refereed to as "VVIQ 2 groups" in the manuscript results.

-   `cluster` is the cognitive style classification obtained from clustering
    OSIVQ scores. It contains three groups: "Visualiser", "Spatialiser", and
    "Verbaliser". It is referred to as "OSIVQ 3 clusters" in the manuscript
    results.

-   `strategy_group` is a classification based on the self-reported strategies
    used by the participants to solve the problems. It focuses on whether
    participants used a visual mental imagery strategy and contains two groups:
    "Visual strategy user" and "No visual strategy". It is not reported in the
    manuscript because it was suggested to us after submission by colleagues at
    a conference.

The same modelling pipeline was therefore applied four times, once for each of
the last four grouping variables.

## The modelling pipeline

We created a small `build_formula()` helper function to write the formula
easily, as we used the same model structure a lot of times:

```{r}
#| label: build-formula

build_formula("accuracy", "group_2")

build_formula("accuracy", "cluster")
```

The models were fitted using the `glmmTMB::glmmTMB()` function (and a little
`set_ranef_prior()` helper function to set a weakly informative prior on the
random effects, which helps with convergence and singularity issues). After
model fit, we checked for potential singularity issues and model performance
using the `get_singularity()` and `get_performance()` functions created for the
occasion (which are convenient wrappers around the *performance* package).

Finally, we tested our hypotheses with marginal contrasts. As stated in the
manuscript:

> Due to the way that variance is partitioned in GLMMs (Rights & Sterba, 2019),
> there does not exist an agreed-upon way to calculate standard effect sizes for
> individual terms such as main effects or interactions in these models. Thus,
> in line with general recommendations on how to report effect sizes (e.g., Pek
> & Flora, 2018), we report and analyse unstandardised effect sizes for post-hoc
> tests in the form of estimated marginal contrasts (i.e. differences in
> model-estimated marginal means, hereinafter denoted Δ), in seconds for RTs or
> as odds ratios for accuracies. To answer our hypotheses, we planned to analyse
> contrasts between groups, contrasts between categories for each group
> separately, and interaction contrasts testing the differences in category
> contrasts between the groups.

This task was performed with the `report_contrast()` function, which is a
wrapper around several functions from the *emmeans* package. We used it to
compute accuracy contrasts between groups, contrasts between categories within
each group, and interaction contrasts (differences in category contrasts between
groups).

Here we go!

# Results

## VVIQ 2 groups

```{r}
#| label: vviq-2-groups

m_acc_vviq_2 <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "group_2"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(55)
  )

# Singularity
m_acc_vviq_2 |> get_singularity()

# Performance
m_acc_vviq_2 |> get_performance() |> knitr::kable(align = "c")
# Group contrasts
m_acc_vviq_2 |> report_contrast(~ group_2) |> knitr::kable()
# Category contrasts within groups
m_acc_vviq_2 |> report_contrast(~ category | group_2) |> knitr::kable()
# Interaction contrasts
m_acc_vviq_2 |> report_contrast(~ category * group_2, interaction = TRUE) |>
  knitr::kable()
```

## VVIQ 3 groups

```{r}
#| label: vviq-3-groups

m_acc_vviq_3 <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "group_3"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(20)
  )

m_acc_vviq_3 |> get_singularity()

m_acc_vviq_3 |> get_performance() |> knitr::kable(align = "c")
m_acc_vviq_3 |> report_contrast(~ group_3) |> knitr::kable()
m_acc_vviq_3 |> report_contrast(~ category | group_3) |> knitr::kable()
m_acc_vviq_3 |> report_contrast(~ category * group_3, interaction = TRUE) |>
  knitr::kable()
```

## OSIVQ 3 clusters

```{r}
#| label: osivq-3-clusters

m_acc_osivq <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "cluster"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(15)
  )

m_acc_osivq  |> get_singularity()

m_acc_osivq  |> get_performance() |> knitr::kable(align = "c")
m_acc_osivq  |> report_contrast(~ cluster) |> knitr::kable()
m_acc_osivq  |> report_contrast(~ category | cluster) |> knitr::kable()
m_acc_osivq  |> report_contrast(~ category * cluster, interaction = TRUE) |>
  knitr::kable()
```

## Strategy groups

```{r}
#| label: strategy-groups

m_acc_strat <-
  glmmTMB::glmmTMB(
    data    = df_expe,
    formula = build_formula("accuracy", "strategy_group"),
    family  = binomial(link = "logit"),
    prior   = set_ranef_prior(55)
  )

m_acc_strat |> get_singularity()

m_acc_strat |> get_performance() |> knitr::kable(align = "c")
m_acc_strat |> report_contrast(~ strategy_group) |> knitr::kable()
m_acc_strat |> report_contrast(~ category | strategy_group) |> knitr::kable()
m_acc_strat |> report_contrast(~ category * strategy_group, interaction = TRUE) |>
  knitr::kable()
```

# Visualisation

The figures showing the distribution of accuracy data displayed in the 
manuscript were created with the `plot_superb_jitter()` function from the
package, which wraps functions from
the *ggplot2* and *superb* packages to create nice visualisations easily. 
A little `add_significance()` helper function was also created to add
significance stars to the plots.

```{r}
#| label: plot-accuracy
#| fig.width: 10
#| fig.height: 4
#| fig.alt: >
#|  A jitter plot showing the mean accuracy in each category for several
#|  different grouping variables.

library(patchwork)
library(superb)

pa1 <-
  plot_superb_jitter(
    df_expe, accuracy, group_2,
    title = "VVIQ 2 groups", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 2,
      y_star = 1.07,
      stars  = "*",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa2 <-
  plot_superb_jitter(
    df_expe, accuracy, group_3,
    title = "VVIQ 3 groups", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 1.07,
      stars  = "*",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa3 <-
  plot_superb_jitter(
    df_expe, accuracy, cluster,
    title = "OSIVQ clusters", y_title = "Mean accuracy",
    base_size = 12,
    plot.background = ggplot2::element_rect(fill = "white", colour = NA),
    axis_rel = 1.2
  ) +
  add_significance(
    size_star = 4,
    tibble::tibble(
      x_star = 3,
      y_star = 1.065,
      stars  = "°",
      x_line = .data$x_star - 0.16,
      x_line_end = .data$x_star + 0.16,
      y_line = 1.05
    )
  )

pa <- pa1 + pa2 + pa3 +
  patchwork::plot_layout(axes = "collect", guides = "collect") &
  ggplot2::theme(legend.position = "bottom")

plot(pa)
```

All done!

--------------------------------------------------------------------------------

```{r session_info}
#| echo: false

sessioninfo::session_info()
```
