---
title: "Creating OSIVQ cognitive style clusters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating OSIVQ cognitive style clusters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
csl: "apa.csl"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This short vignette details how the OSIVQ clusters were created and named, and 
how they relate to the VVIQ-based groups.

```{r setup}
library(aphantasiaReasoningViie)
```

First, let's get the cleaned, analysis-ready data (see 
`vignette("preparing_data")` for details). 

```{r data}
df_survey  <- get_clean_data()$df_survey
```

The OSIVQ clusters were created using a consensus of three clustering algorithms
(GMM, PAM and C-Means) applied to the three OSIVQ sub-scales (Object, Spatial and
Verbal). A function was created for this specific task, `cluster_osivq()`, which
uses the `diceR::dice()` function from the *diceR* package under the hood.
Following the methodology proposed by @delem2025uncovering, we searched for 
three "visualiser", "spatialiser" and "verbaliser" clusters based on their most
"dominant" OSIVQ sub-scale score.

However, the clustering algorithm does not name the clusters, so we needed to
check their properties to assign meaningful names. The function
`add_named_clusters()` adds the cluster assignments to the data frame, by default
with generic names. `summarise_clustering()` then provides a summary of the
clusters' properties, which allowed us to choose meaningful names for the three
clusters produced.

```{r cluster_and_check}
# Clustering OSIVQ data
clustering <- cluster_osivq(df_survey)

# Checking cluster properties to define names for each cluster
df_survey |>
  add_named_clusters(clustering) |>
  summarise_clustering()
```

We saw here that cluster 1 was the spatialiser one, cluster 2 was the visualiser
and cluster 3 the verbaliser. We can now use `add_named_clusters()` again with 
its optional arguments to add these labels, reorder them, choose factor levels
and add contrasts for planned comparisons between the visualiser cluster and the
other two.

```{r labelling}
# Adding named clusters to the survey data
df_survey <- add_named_clusters(
  df_survey, clustering,
  names  = c("Spatialiser", "Visualiser", "Verbaliser"),
  levels = c("Visualiser", "Spatialiser", "Verbaliser"),
  contrasts = c("_visualiser", "_spatialiser", "_verbaliser"),
  base = 1
)

# Let's check the cluster properties again
contrasts(df_survey$cluster)
summarise_clustering(df_survey)
```

`cluster_osivq()` uses a seed, so the resulting clusters will always be the 
same for a given seed and data set. Thus, we were able to use this short 
clustering pipeline in all subsequent analyses to create our "cluster" 
variable and add it to the experimental data set for modelling.
