---
title: "Power analysis by simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Power analysis by simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "references.bib"
csl: "apa.csl"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette dives into the details of the power analysis by simulation that 
is briefly presented in the manuscript (preprint [here](https://doi.org/10.31234/osf.io/vsjtb_v1)). We describe the rationale behind
this approach, how we built the generative model used to simulate data, how we
fitted models on these simulations, and how we estimated the power of our model
to detect the effect of interest.

# Rationale

We analysed sample and effect sizes before data collection using a simulation
approach, where power is defined as the proportion of simulations where a model
detects an existing effect that we simulated. Instead of choosing a fixed sample
size, this approach allowed us to have a full picture of the power of our model
across a range of sample sizes and effect sizes. It replaced the question "*What
sample size do we need to detect an effect of a given (arbitrary) size?*" with
"*Given the sample we managed to recruit, what is the smallest effect size we
could detect with good power?*" which allowed much more flexibility in data
collection. This was especially useful for us because aphantasia and
hyperphantasia are rare phenomena [see
@wrightInternationalEstimatePrevalence2024], and we had no way of knowing
beforehand how many participants we would be able to recruit in each group.

# Building the generative model

The process starts with finding a way to generate synthetic data (said
"simulations") that "look like" the experimental data we expected to collect. We
based our simulations on a model that we hypothesised to govern the effect of
our variables of interest on the dependent variables, and distributional
assumptions about these dependent variables. Below, we describe this model and
the previous literature from which we derived the estimations of the effects and
the distributions of the observed variables.

## Structure

We designed our model before data collection based on our hypotheses and
variables. This is the model that was be used as the data-generating process for
the simulations. The model we used is a generalised linear mixed-effects model
(GLMM) with the following structure, in R syntax:

``` r
dependent_variable ~ grouping * category + (category | id) + (1 | problem)
```

Where the dependent variable can be accuracy or response times (RTs).

-   The `grouping` can be our basic VVIQ groups (of varying precision depending
    on the definition, with 2, 3 or 4 groups), or OSIVQ cognitive style groups
    as proposed by @delem2025uncovering (in aphantasia literature) or
    @gazzoIndividualDifferencesImagery2013 (in reasoning literature).

-   The `category` is the problem category (visual, spatial or control).

-   The interaction between grouping and category is a third fixed factor
    (implied by the `*`).

-   The random effects are the intercepts and slopes by category for each
    participant (`category | id`) and an intercept per problem.

The parameters for each variable can be made explicit by writing the model
formally as such:

\begin{align*}
DV = \ & \beta_0 + \tau_0 + \\
  & (\beta_{grouping}) \times grouping \ + \\
  & (\beta_{category} + \tau_{category}) \times category \ + \\
  & (\beta_{grouping:category} + \tau_{grouping:category}) \times grouping \times category \ + \\
  & \epsilon
\end{align*}

Where:

-   $DV$ is the dependent variable (RTs or accuracy).
-   $\beta_0$ is the global intercept (the mean of the dependent variable).
-   $\tau_0$ is the random intercept for each participant (the variation of the
    global intercept per participant).
-   $\beta_{grouping}$ is the fixed effect of the grouping variable (the
    difference in the dependent variable between the groups).
-   $\beta_{category}$ is the fixed effect of the category variable (the
    difference in the dependent variable between the categories).
-   $\tau_{category}$ is the random effect of the category (the variation of the
    category effect per participant).
-   $\beta_{grouping:category}$ is the fixed effect of the interaction between
    the grouping and category variables (the difference in the dependent
    variable between the groups for each category).
-   $\tau_{grouping:category}$ is the random effect of the interaction between
    the grouping and category variables (the variation of the interaction effect
    per participant).
-   $\epsilon$ is the residual error (the variation of the dependent variable
    that is not explained by the model).

There are actually two $\beta_{category}$, one for each category (minus one, for
the reference level), and as many $\beta_{grouping:category}$ as there are
combinations of categories and groups. In practice, for our purposes, we are
only interested in the difference between the visual category and the others,
that we called the $\beta_{vis}$ coefficient, and the difference in this
difference (the interaction) for a specific group (aphantasia), that we called
the $\beta_{aph-vis}$ coefficient. Thus, for our simulations, we set
$\beta_{grouping}$, $\beta_{spatial}$ (the effect of the
spatial category) and $\tau_{spatial}$ (the associated slope) and all other
interaction coefficients to 0, respectively because we had no hypothesis or
previous data on the group effect, the effect of the problems, the
spatial effect, or the variation in spatial performance.

Given this theoretical model and our assumptions, we tried to choose constant
values for the remaining parameters that were not analysed ($\beta_{0}$,
$\tau_{0}$ and $\tau_{vis}$) and ranges of values for the parameters of interest
in our power analysis ($\beta_{vis}$ and $\beta_{aph-vis}$) based on previous
literature.

## Literature reference for parameter values

We based the power analysis on the Visual Imagery Impedance Effect (VIIE) on
RTs, which were the most robust and well-documented we could find. More
precisely, we used the very well-described data and model from
@tseVisualImpedanceEffect2017. They had a study and paradigm very similar to
ours, so their statistics were fairly simple to adapt to determine good
simulation parameters. The main difference is that they used continuous and
discontinuous problems, while we used only semi-continuous problems, so we chose
to average the values of the two types of problems from their study.

Their statistics were:

-   Control category mean RT: 14.641 seconds

-   Visual category mean RT: 16.388 seconds

-   Spatial category mean RT: 14.197 seconds

Their Generalized Estimating Equations also showed:

-   An intercept of 12.051 seconds, 95% CI = [10.739, 13.363]

To which we can add the $\beta$ coefficient of the 4/5-terms (they are the
same):

-   $\beta$ 4T: 2.482 seconds, 95% CI = [1.607, 3.356]

-   Finally, the visual category $\beta$: 2.633 seconds, 95% CI = [1.121, 4.146]

Thus, for our simulations, **we aimed for the following parameters:**

-   A global mean $\beta_0$ around 14.5 seconds.

-   Variations between 12.5 and 16.5 seconds, which we translated as varying
    intercepts per subject $\tau_0$ ranging between -2 and 2 seconds.

-   A fixed effect of the visual category $\beta_{vis}$ around 2.6 seconds.

-   Variations in this fixed effect between 1.1 and 4.1 seconds, which we
    translated as varying slopes per subject $\tau_{vis}$ ranging between -1.5
    and 1.5 seconds.

## Finding the simulation parameters

We hypothesised that the aphantasia group would not show a visual effect while
the typical group would, a pattern that would result in an interaction between
the visual category and the aphantasia group. The visual effect in the typical
group would result in a positive $\beta_{vis}$ coefficient, while the
aforementioned interaction would result in a negative $\beta_{aph-vis}$
interaction coefficient that nullifies the visual effect for the aphantasia
group only.

We chose to simulate the response times with a shifted log-normal distribution
(generated with the package `brms`), as these distributions reproduce
particularly well those of RTs (see [this amazing interactive
illustration](https://lindeloev.shinyapps.io/shiny-rt/)). We searched manually
for the parameters of this distribution and our model coefficients that allowed
to reach the desired statistics. The data simulation using the generative model
is wrapped in the `simulate_rt_data()` function. The function has arguments to
fine-tune all the distributional parameters and model coefficients discussed
above. We searched for the appropriate values by trial-and-error using the
simple code below:

```{r setup}
library(aphantasiaReasoningViie)
```

```{r}
#| label: searching-correct-parameters

df_test <- 
  simulate_rt_data(
    n_subj_per_group = 100, 
    # Parameters of the shifted log-normal distribution (from brms)
    meanlog = 2.1,
    sdlog   = 0.55, 
    shift   = 5,
    # Varying intercept by-subject
    tau_0 = 0.9,
    # Visual category effect
    beta_vis = 2.35,
    # Varying visual effect by-subject
    tau_vis  = 0.75,
    # Aphantasia group x visual interaction
    beta_aph_vis = -2.35
  )

df_test |> 
  dplyr::group_by(category, group) |>  
  dplyr::reframe(
    mean   = mean(rt_total),     
    median = median(rt_total),  
    min    = min(rt_total), # To test the shift     
    max    = max(rt_total)  # To test the sdlog dispersion effect
  ) 
```

Using the argument values above, we see that we managed to reach statistics
close to the ones we wanted, while simulating an interaction effect between the
groups in the visual category. We found that:

-   A mean of 2.1, SD of 0.55 seconds and a shift (non-decision time) of 5
    seconds on the shifted log-normal distribution allowed to reach the desired
    $\beta_0$ of 14.5 seconds reliably (testing with no other source of
    variation, i.e., all other parameters at 0).

-   A varying intercept $\tau_0$ with a SD of 0.9 allowed to obtain
    distributions of the varying RTs ranging between -2 and 2 seconds on
    average, as expected.

-   A visual effect $\beta_{vis}$ of 2.35 allowed to reproduce the visual
    category means observed in the study reliably.

-   A varying visual effect $\tau_{vis}$ with a SD of 0.75 allowed to obtain
    distributions of the varying visual RTs ranging between -1.5 and 1.5 seconds
    on average, as expected.

-   An interaction parameter $\beta_{aph-vis}$ set as minus the visual effect
    allowed to nullify the visual effect in the aphantasia group. If the effect
    size gets small, a slight multiplier (e.g., 1.5) might be necessary to
    nullify the effect.

We can visualise the distribution of these RTs using the
`plot_superb_raincloud()` function from the package, which wraps functions from
the *ggplot2* and *superb* packages to create nice visualisations easily:

```{r}
#| label: plot-distribution-simulated-rt
#| fig.width: 6
#| fig.height: 4
#| fig.alt: >
#|  Raincloud plot showing the distribution of simulated response times
#|  across the visual, spatial and control categories and the
#|  aphantasia and typical groups.

library(superb)

plot_superb_raincloud(
  df_test, rt_total, group,
  title = "Simulated response times per group and category",
  y_title = "Mean total RT (s)",
  base_size = 12,
  plot.background = ggplot2::element_rect(fill = "white"),
  axis_rel_x = 1.2,
  exp_add_right = 0.7
  )
```

The slowdown in the visual category for the typical group appears very clearly.
Now the whole purpose of the power analyses will be to play around with this RT
difference to see how well and how often our model is able to detect it.

All set! Now we can move on to the actual power analysis.

# Fitting models on a lot of simulations

## Modelling and testing the contrast of interest

For our final analyses, we fitted frequentist GLMMs with Gamma distributions
using the `glmmTMB` package. For the sake of computation speed, and because we
did not simulate any variance tied to the problem types, we simplified the
formula for the power analyses by removing the second random term. *Note the use
of the `set_ranef_prior()` helper function, which helps setting regularizing
priors on the random effects' SD to alleviate convergence issues*.

Let's see how this model performs on our test data. We'll use the
`report_contrast()` function from the package to get the contrasts of interest
easily.

```{r}
#| label: fitting-model-test-data

model <- 
  glmmTMB::glmmTMB(
    rt_total ~ group * category + (category | id),
    data = df_test,
    family = Gamma(link = "identity"),
    prior   = set_ranef_prior(70)
  )

report_contrast(model, ~ category | group) |> knitr::kable()
```

We see that the model managed to detect the visual effect in the typical group,
and also does a pretty decent estimation of its value (we simulated a 2.35s
effect, with some random variance added). How about the interaction contrasts?

```{r}
#| label: interaction-contrast-test-data

report_contrast(model, ~ group * category, interaction = TRUE) |> knitr::kable()
```

As expected, the interaction is also detected, and its value reflects the
difference in the visual effect between the groups.

We used the *p*-value of the Control-Visual interaction contrast as the
criterion to determine whether the model was successful or not in detecting the
effect. Thus, a single loop of the power analysis looks like this:

```{r}
#| label: single-simulation-loop

# Choose sample and effect sizes to test
n_subj_per_group <- 40
beta_vis <- 1

# Simulate data
df <- 
  simulate_rt_data(
    n_subj_per_group = n_subj_per_group, 
    beta_vis = beta_vis,          # Visual category effect
    beta_aph_vis = -1 * beta_vis, # group x visual interaction
    seed = 1234
  )

# Fit the model on the simulated data
model <-
  glmmTMB::glmmTMB(
    rt_total ~ group * category + (category | id),
    data = df,
    family = Gamma(link = "identity"),
    prior   = set_ranef_prior()
  ) |> suppressMessages() |> suppressWarnings()

# Get the p-value of the control-visual interaction contrast
contrasts <- report_contrast(model, ~ group * category, interaction = TRUE)
p <- contrasts$p.value[2]

print(p)
```

We wrapped this loop in the `simulate_rt_test()` function:

```{r}
#| label: simulate-rt-test-function

simulate_rt_test(n_subj_per_group = 40, beta_vis = 1, seed = 1234)
```

Now time for power! `run_power_analysis()` takes a range of sample sizes, a
range of effect sizes ($\beta_{vis}$), and a number of simulations per
sample/effect size combination. It then runs the `simulate_rt_test()` function
on all these combinations, and returns the results of the power analysis. This
function is interactive (it gives info on what's going to be computed and waits
for user input) so it does not run in this notebook, but feel free to try it
out!

This is the code that was used to run the power analysis that is presented in
the manuscript:

```{r}
#| label: run-power-analyses-example
#| eval: false

power_results <- 
  run_power_analysis(
    n_min        = 10,
    n_max        = 200,
    n_step       = 10,
    beta_vis_min = 0.5,
    beta_vis_max = 2.5,
    beta_step    = 0.1,
    n_simulations = 350
  )
```

... This is a total of 147,000 models fitted! It took 16 hours and 44 minutes
(although the time can vary a lot depending on the computer), so of course we
saved the results to a file for later use. These results are provided as package
data in the `power_results` dataset, so they can be accessed easily:

```{r}
#| label: check-power-results

power_results |>
    dplyr::group_by(n_subj_per_group, beta_vis) |>
    dplyr::reframe(power = sum(p_value <= 0.05) / dplyr::n()) 
```

The table above shows the estimated power for each combination of sample size
and effect size. The package also provides a `plot_power()` function to
visualise these results with nice curves:

```{r}
#| label: plot-power-results
#| fig.width: 8
#| fig.height: 5
#| fig.alt: >
#|  Line plot showing the estimated power to detect the visual effect
#|  across a range of sample sizes and effect sizes.

plot_power(
  power_results, 
  base_size = 12,
  plot.background = ggplot2::element_rect(fill = "white")
)
```

# Conclusion

Mission accomplished! The power analysis by simulation allowed us to have a full
picture of the power of our model across a range of sample sizes and effect
sizes. It made the interpretation of the results of our analyses on real data
much easier and gave straightforward directions on the way to go to gather
evidence in favour or against the existence of the VIIE.

--------------------------------------------------------------------------------

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```

## References
